---
title: "How SQL DB server works"
author: "Ricardo Rodrigo Basa"
date: "2024-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, include=FALSE}
library(dplyr)
library(DBI)
library(glue)
library(DT)
```

In the previous chapter, we set up our sandbox table `sample_table` with 4 columns and 50
million rows. Now, we will execute a few queries to understand how SQL databases retrieve
the rows we request.

## Connect to the DB

```{r connect_to_db}
conn <- dbConnect(
  RMariaDB::MariaDB(),
  host = Sys.getenv("DB_HOST"),
  port = 3306,
  username = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASSWORD"),
  dbname = Sys.getenv("DB_SCHEMA")
)
```

## Initial exploration

Let's take a look at what we are dealing with. We know we have `entry_date` in the table sorted
in ascending order.

```{r date_limits}
dbGetQuery(conn, "SELECT MIN(entry_date), MAX(entry_date) FROM sample_table")
```

## Start MySQL profiling

```{r set_profiling, echo=T, results='hide'}
dbExecute(conn, "SET @@profiling = 1")
```

## Query 1: Retrieve just the first day

We know that the first day we have data for is at the top of the table.

```{r query_1}
result <- dbGetQuery(conn, "SELECT * FROM sample_table WHERE entry_date = '1973-01-01'")
query_1_count <- result %>% nrow()
datatable(result)
```
We have `r query_1_count` rows in our result set. We can verify that we only retrieved 
1973-01-01, our first entry date. And we see that the primary key `id` begins at `1`.

To check how long the query took, we execute the `SHOW PROFILES` command.

```{r profiles_1}
profiles <- dbGetQuery(conn, "SHOW PROFILES")
profiles %>%
  slice_tail(n = 3)
```

It takes around 17 seconds.

*The `DBI` library performed `commit` DB commands after the query, and again, after the `SHOW
PROFILES` command.*

If you are using a GUI client to connect to your DB, it may show a time for the query to
execute. In MySQL Workbench, this would be shown as *Duration/Fetch*. And for this query,
MySQL Workbench shows `0.029 sec / 17.224 sec`. The sum of these is roughly the same as the 
duration shown in the profiles result.

```{r profile_1}
this_profile <- (profiles %>% slice_tail(n = 3))[1, ]
this_query_id <- this_profile %>% pull(Query_ID)

dbGetQuery(conn, glue("SHOW PROFILE FOR QUERY {this_query_id}"))
```

The majority of the time is spent in `executing`.

## Query 2: Retrieve just the last day

Let's do the same for the last day we have, 2023-12-31. We know that the last day of the data
we have is at the end of the table.

```{r query_2}
result <- dbGetQuery(conn, "SELECT * FROM sample_table WHERE entry_date = '2023-12-31'")
query_2_count <- result %>% nrow()
datatable(result)
```

```{r profiles_2}
profiles <- dbGetQuery(conn, "SHOW PROFILES")
profiles %>%
  slice_tail(n = 3)
```

```{r profile_2}
this_profile <- (profiles %>% slice_tail(n = 3))[1, ]
this_query_id <- this_profile %>% pull(Query_ID)

dbGetQuery(conn, glue("SHOW PROFILE FOR QUERY {this_query_id}"))
```

We get the same time. Roughly 17 seconds.

Let us investigate further.

## Query 3: Retrieve the first day but indicate how many rows to retrieve

We know our 1973-01-01 query resulted in `r query_1_count` rows. Let's add that `LIMIT`
to our query.

```{r query_3}
result <- dbGetQuery(conn, 
                     glue("SELECT * FROM sample_table WHERE entry_date = '1973-01-01' 
                          LIMIT {query_1_count}"))

this_profile <- dbGetQuery(conn, "SHOW PROFILES") %>%
  slice_tail(n = 3) %>%
  slice_head(n = 1)
this_query_id <- this_profile %>% pull(Query_ID)

this_profile
dbGetQuery(conn, glue("SHOW PROFILE FOR QUERY {this_query_id}"))
```

Now, our query is executed in less than 0.005 seconds.

## Query 4: Retrieve the last day but indicate how many rows to retrieve

We also know that the 2023-12-31 query resulted in `r query_2_count` rows. Let's do
the same.

```{r query_4}
result <- dbGetQuery(conn, glue("SELECT * FROM sample_table WHERE entry_date = '2023-12-31'
                                LIMIT {query_2_count}"))

this_profile <- dbGetQuery(conn, "SHOW PROFILES") %>%
  slice_tail(n = 3) %>%
  slice_head(n = 1)
this_query_id <- this_profile %>% pull(Query_ID)

this_profile
dbGetQuery(conn, glue("SHOW PROFILE FOR QUERY {this_query_id}"))
```

We are back to around 17 seconds.

## Query 5: Retrieve first day but request for `r query_1_count + 1` rows

One last test:

```{r query_5}
result <- dbGetQuery(conn, 
                     glue("SELECT * FROM sample_table WHERE entry_date = '1973-01-01' 
                          LIMIT {query_1_count + 1}"))

this_profile <- dbGetQuery(conn, "SHOW PROFILES") %>%
  slice_tail(n = 3) %>%
  slice_head(n = 1)
this_query_id <- this_profile %>% pull(Query_ID)

this_profile
dbGetQuery(conn, glue("SHOW PROFILE FOR QUERY {this_query_id}"))
```

Our query to retrieve `r query_1_count + 1` rows takes 17 seconds. 

## Questions

1. Can you explain why queries for the first day (Q1) and the last day (Q2) take the same amount of time?
2. Can you explain why the query for the first day `LIMIT` `r query_1_count` (Q3) is almost instantaneous, but `LIMIT` `r query_1_count + 1` (Q5) takes a very long time?
3. Can you explain why the query for the last day `LIMIT` `r query_2_count` (Q4) takes the same time as Q1 and Q2?

```{r discconect_db, echo=T, results='hide'}
dbExecute(conn, "SET @@profiling = 0")
dbDisconnect(conn)
```
